### 2023-03-02 calculus.tex
Edit calculus section and update rules

\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \\
    \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV}
                       \mid           \ENumMV \mid \EPlus{\EMV}{\EMV}
                       \mid           \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
                       \mid           \EEHole \\
    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV}
                       \mid           \ECNumMV \mid \ECPlus{\ECMV}{\ECMV}
                       \mid           \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV}
                       \mid           \EEHole \\
             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
  \end{array}\]
  \caption{Syntax of an extended gradually typed lambda calculus.}
  \label{fig:calculus-syntax}
\end{figure}

Types classify expressions by a \emph{bidirectional type system}, which employs two mutually defined
judgments. \emph{Type synthesis}, written $\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under
the typing context $\ctx$, the expression $e$ synthesizes or infers the type $\TMV$. \emph{Type
analysis}, written $\ctxAnaType{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear
where an expression of type $\TMV$ is expected. We may give these typing semantics traditionally,
without considering error localization and recovery.

@@

\Cref{fig:calculus-syntax} furthermore gives the syntax for marked expressions (as $\ECMName$),
which replicates all the forms of unmarked expressions. There are additionally three kinds of
\emph{error marks}, which represent the different type errors that may arise.
%
\begin{itemize}
  \item $\ECUnbound{x}$, the \emph{unbound variable mark}, indicates the use of an unbound variable
    $x$, as in \cref{fig:calculus-examples-unbound}.

  \item $\ECInconType{\ECMV}$, the \emph{inconsistent type mark}, indicates that the type of an
    expression does not match an expected type.

  \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, the \emph{inconsistent branches mark}, indicates
    that the branches of a conditional have inconsistent types.
\end{itemize}
%
Throughout the remainder of this section, as we consider the formulation of marking, we will
motivate and give precise semantics for these error marks. Note that this is a non-exhaustive list;
depending on the syntactic forms and typing features found in the language, other kinds of marks may
be necessary. Indeed, in \cref{sec:calculus-poly}, we consider the kinds of marks that may be
required in an language with explicit polymorphism.

The marked language possesses its own typing semantics, also formulated bidirectionally. We write
$\ctxSynType{\ctx}{\ECMV}{\TMV}$ for synthesis and $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ for analysis.
Finally, marking is bidirectional as well. The synthetic marking judgment
$\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that, under the context $\Gamma$, the
unmarked expression $\EMV$ is ``marked into'' the marked expression $\ECMV$, which synthesizes type
$\TMV$. Analogously, the analytic marking judgment $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
states that $\EMV$ is marked into $\ECMV$, which analyzes against $\TMV$.

@@

Furthermore, marking should preserve the syntactic structure of programs, modulo error marks. Thus,
we define a notion of \emph{mark erasure}, whose definition is given
\cref{fig:calculus-mark-erasure}, which converts marked expressions back into unmarked expression by
removing error marks. Then, \cref{thm:calculus-marking-well-formedness} provides a
\emph{well-formedness} criterion marking.
%
\begin{theorem}[name=Marking Well-Formedness] \
  \label{thm:calculus-marking-well-formedness}
  \begin{enumerate}
    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxSynType{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
  \end{enumerate}
  % TODO split this into two theorems?
  % TODO not really sure how to intuitively describe the part about types matching up
\end{theorem}
%
In other words, if an unmarked expression is marked into some marked expression, erasing error marks
from the marked expression gives the original unmarked expression. We also ensure that marking gives
well-typed marked expressions; indeed, their types should match the types synthesized or analyzed
against in marking. % TODO more intuition for this theorem

@@

In the remainder of this subsection, we consider the different type errors that may arise in the
unmarked language; in this way, the typing semantics of the unmarked language guide the construction
of the marking process.

@@

% TODO need to say something about subsumption and how it will be handled later.
To start, let us consider the simple case of numbers. Because of subsumption, we need only define a
synthesis rule for unmarked numbers, in which numbers synthesize the type $\TNum$:
%
\begin{mathpar}
  \judgment{ }{
    \ctxSynType{\ctx}{\ENumMV}{\TNum}
  }{USNum}
\end{mathpar}

Into what should numbers be marked? Quite straightforwardly, we simply give the same number as a
marked expression, synthesizing again $\TNum$. No type errors may occur in a single number, so we
only need the following rules for marking unmarked numbers and typing marked numbers:
%
\begin{mathpar}
  \judgment{
  }{
    \ctxSynFixedInto{\ctx}{\ENumMV}{\ECNumMV}{\TNum}
  }{MKSNum}

  \judgment{ }{
    \ctxSynType{\ctx}{\ECNumMV}{\TNum}
  }{MSNum}
\end{mathpar}

In addition expressions, the type of both operands should be $\TNum$. In a bidirectional system, we
thus analyze the operands against $\TNum$, giving the following rule:
%
\begin{mathpar}
  \judgment{
    \ctxAnaType{\ctx}{\EMV_1}{\TNum} \\
    \ctxAnaType{\ctx}{\EMV_2}{\TNum}
  }{
    \ctxSynType{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\TNum}
  }{USPlus}
\end{mathpar}

The marking rule mirrors \textsc{USPlus} closely. Since a known, expected type for each operand is
known, we shift the ``blame'' for any type errors to them. Hence, we recursively mark each operand
in analytic mode and rebuild the marked addition expression. The marked typing rule then mirrors
\textsc{USPlus} exactly.
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TNum} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TNum}
  }{
    \ctxSynFixedInto{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{MKSPlus}

  \judgment{
    \ctxAnaType{\ctx}{\ECMV_1}{\TNum} \\\\
    \ctxAnaType{\ctx}{\ECMV_2}{\TNum}
  }{
    \ctxSynType{\ctx}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{MSPlus}
\end{mathpar}

Let us now consider the more interesting case of variables. The typing of unmarked variables in the
gradually typed lambda calculus is standard. Again, because of subsumption, there is only a
synthetic rule:
\[%
  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynType{\ctx}{x}{\TMV}
  }{USVar}
\]%
If $x$ is bound to a type in the typing context, it synthesizes that type. Straightforwardly,
marking converts the unmarked $x$ into the same marked $x$ via the following marking and typing
rules:
%
\begin{mathpar}
  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{x}{x}{\TMV}
  }{MKSVar}

  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynType{\ctx}{x}{\TMV}
  }{MSVar}
\end{mathpar}
% TODO more intuition on why this makes sense?

But what happens a variable, such as $y$ in \cref{fig:calculus-examples-unbound}, is \emph{not}
bound, i.e. that the premise fails? Then, \textsc{USVar} does not apply, and the typing semantics
say nothing about such expressions. When devising marking, to satisfy totality, such a possibility
motivates an \emph{unbound variable mark} $\ECUnbound{x}$, accompanied by the following rules:
%
\begin{mathpar}
  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynFixedInto{\ctx}{x}{\ECUnbound{x}}{\TUnknown}
  }{MKSUnbound}

  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynType{\ctx}{\ECUnbound{x}}{\TUnknown}
  }{MSUnbound}
\end{mathpar}
%
\textsc{MKSUnbound} states that an variable unbound in the context is marked into that variable
inside an unbound variable mark. \textsc{MSUnbound} then gives the typing of unbound variable marks.
Since nothing may be said about their types, we synthesize the unknown type, which allows type
checking to continue---instead of simply failing---and the unbound variable to be used in any
expression. % TODO clarify why this is correct

There are explicit synthesis and analysis rules for unmarked lambda abstractions:
%
\begin{mathpar}
  \judgment{
    \ctxSynType{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{USLam}

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\
    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxAnaType{\ctx}{\ECLam{x}{\TMV}{\EMV}}{\TMV_3}
  }{UALam}
\end{mathpar}
%
The synthesis rule is standard. In the premises of \textsc{UALam}, we use the notion of
\emph{type consistency} from gradual type theory, which defines a reflexive and symmetric (but not
transitive) relation between types, writing $\consistent{\TMV_1}{\TMV_2}$ to mean that $\TMV_1$ is
consistent with $\TMV_2$. This replaces the notion of \emph{type equality} to relate the unknown
type to all other types. The analysis rule also employs the judgment
$\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$, which establishes that $\TMV$ is a \emph{matched arrow type},
i.e. it may be considered an arrow type. This notion is purely a technical mechanism to avoid
duplication of rules related to arrow types. \Cref{fig:calculus-consistency-matched-arrow} gives the
rules for both judgments.

@@

    \judgment{
      \consistent{\TMV_1}{\TMV_1'} \\
      \consistent{\TMV_2}{\TMV_2'} \\
    }{
      \consistent{\TArrow{\TMV_1}{\TMV_2}}{\TArrow{\TMV_1'}{\TMV_2'}}
    }{TCArr} \\
  \end{mathpar}

  \judgbox{\ensuremath{\matchedArrow{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched arrow type $\TArrow{\TMV_1}{\TMV_2}$
  %
  \begin{mathpar}
    \judgment{ }{
      \matchedArrow{\TUnknown}{\TUnknown}{\TUnknown}
    }{TMAHole}

    \judgment{ }{
      \matchedArrow{\TArrow{\TMV}{\TMV}}{\TMV}{\TMV}
    }{TMAArr}
  \end{mathpar}
  %
  \caption{Type consistency and matched arrow types.}
  \label{fig:calculus-consistency-matched-arrow}
\end{figure}

@@

  \judgment{
    \ctxSynType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
  }{
    \ctxSynType{\ctx}{\ECLam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{MSLam}
\end{mathpar}
%
Similarly, we construct corresponding analytic rules:
%
\begin{mathpar}
  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\\\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MKALam1}

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\\\
    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
  }{
    \ctxAnaType{\ctx}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MALam1}
\end{mathpar}
%
However, recalling totality, we look again to the premises of \textsc{UALam} to see what type errors
may arise. First, what if $\TMV_3$ is not a matched arrow type? This would be the case if it were
$\TNum$, for example, and the lambda abstraction would be inconsistent with the expected type. Thus,
to the syntax we add an \emph{inconsistent type mark} $\ECInconType{\ECMV}$, which indicates that
the expression does not synthesize a type consistent with the expected type. We then have the
following rules, in which the lambda abstraction is placed inside the inconsistent type mark.
%
\begin{mathpar}
  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TUnknown}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
  }{MKALam2}

  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TUnknown}
  }{
    \ctxAnaType{\ctx}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
  }{MALam2}
\end{mathpar}
%
Though no expected output type is known, we still need to check and mark the body; it is thus
analyzed against the unknown type in \textsc{MKALam2}.

>>>>

\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \mid \TProd{\TMV}{\TMV} \\
    \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV} \mid \ELet{x}{\EMV}{\EMV}
                       \mid           \ENumMV \mid \EPlus{\EMV}{\EMV} \\
             &       & \mid         & \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
                       \mid           \EPair{\EMV}{\EMV}
                       \mid           \EProjL{\EMV} \mid \EProjR{\EMV}
                       \mid           \EEHole \\
    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV} \mid \ECLet{x}{\ECMV}{\ECMV}
                       \mid           \ECNumMV \mid \ECPlus{\ECMV}{\ECMV} \\
             &       & \mid         & \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV}
                       \mid           \ECPair{\ECMV}{\ECMV} \mid \ECProjL{\ECMV} \mid \ECProjR{\ECMV}
                       \mid           \ECEHole \\
             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV} \mid \ECInconAsc{\ECMV} \\
             &       & \mid         & \ECSynNonMatchedArrow{\ECMV} \mid \ECAnaNonMatchedArrow{\ECMV}
                       \mid           \ECSynNonMatchedProd{\ECMV} \mid \ECAnaNonMatchedProd{\ECMV}
  \end{array}\]
  \caption{Syntax of an extended gradually typed lambda calculus.}
  \label{fig:calculus-syntax}
\end{figure}

Types classify expressions by a \emph{bidirectional type system}, which employs two mutually defined
judgments. \emph{Type synthesis}, written $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$, establishes that, under
the typing context $\ctx$, the expression $e$ synthesizes or infers the type $\TMV$. \emph{Type
analysis}, written $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear
where an expression of type $\TMV$ is expected. We may give these typing semantics traditionally,
without considering error localization and recovery.

@@

\Cref{fig:calculus-syntax} furthermore gives the syntax for marked expressions (as $\ECMName$),
which replicates all the forms of unmarked expressions. There are additionally several kinds of
\emph{error marks}, which represent the different type errors that may arise. Intuitively, they each
denote a different error message that might be shown in an editor or emitted by a compiler.
%
% \begin{itemize}
  % \item $\ECUnbound{x}$, the \emph{unbound variable mark}, indicates the use of an unbound variable
    % $x$, as in \cref{fig:calculus-examples-unbound}.
%
  % \item $\ECInconType{\ECMV}$, the \emph{inconsistent type mark}, indicates that the type of an
    % expression does not match an expected type.
%
  % \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, the \emph{inconsistent branches mark}, indicates
    % that the branches of a conditional have mismatched types.
%
  % \item $\ECLamInconAsc{x}{\TMV}{\ECMV}$, the \emph{inconsistent ascription mark}, indicates that
    % the type ascription $\TMV$ is mismatches with an expected type.
%
  % \item $\ECSynNonMatchedArrow{\ECMV}$
% \end{itemize}
%
Throughout the remainder of this section, as we formulate marking for the gradually typed lambda
calculus, we will motivate and give precise semantics for these error marks. Note that this is a
non-exhaustive list; depending on the syntactic forms and typing features found in the language,
other kinds of marks may be necessary. Indeed, in \cref{sec:calculus-poly}, we consider the kinds of
marks that may be required in an language with explicit polymorphism.

The marked language possesses its own typing semantics, also formulated bidirectionally. We write
$\ctxSynTypeM{\ctx}{\ECMV}{\TMV}$ for synthesis and $\ctxAnaTypeM{\ctx}{\ECMV}{\TMV}$ for analysis.
Finally, marking is bidirectional as well. The synthetic marking judgment
$\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that, under the context $\Gamma$, the
unmarked expression $\EMV$ is ``marked into'' the marked expression $\ECMV$, which synthesizes type
$\TMV$. Analogously, the analytic marking judgment $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
states that $\EMV$ is marked into $\ECMV$, which analyzes against $\TMV$.

@@

Furthermore, marking should preserve the syntactic structure of programs, modulo error marks. Thus,
we define a notion of \emph{mark erasure}, whose definition is given
\cref{fig:calculus-mark-erasure}, which converts marked expressions back into unmarked expression by
removing error marks. Then, \cref{thm:calculus-marking-well-formedness} provides a
\emph{well-formedness} criterion marking.
%
\begin{theorem}[name=Marking Well-Formedness] \
  \label{thm:calculus-marking-well-formedness}
  \begin{enumerate}
    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxSynTypeM{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxAnaTypeM{\ctx}{\ECMV}{\TMV}$ and
      $\erasesTo{\ECMV}{\EMV}$.
  \end{enumerate}
  % TODO split this into two theorems?
  % TODO not really sure how to intuitively describe the part about types matching up
\end{theorem}
%
In other words, marking and then erasing error marks from the result gives the original unmarked
expression. We also ensure that marking gives well-typed marked expressions; indeed, their types
should match the types synthesized or analyzed against in marking.
% TODO more intuition for this theorem

@@

In the remainder of this subsection, we consider the different type errors that may arise in the
unmarked language while keeping in mind these two theorems; in this way, the typing semantics of the
unmarked language guide the construction of the marking process.

@@

To start, let us consider the simple case of numbers. Because of \emph{subsumption}, which we
discuss next, we need only define a synthesis rule for unmarked numbers, in which numbers synthesize
the type $\TNum$:
%
\begin{mathpar}
  \judgment{ }{
    \ctxSynTypeU{\ctx}{\ENumMV}{\TNum}
  }{USNum}
\end{mathpar}

How should numbers be marked? Quite straightforwardly, we simply give the same number as a marked
expression, synthesizing again $\TNum$. No type errors may occur in a single number, so we only need
the following rules for marking numbers and typing the results: 
%
\begin{mathpar}
  \judgment{
  }{
    \ctxSynFixedInto{\ctx}{\ENumMV}{\ECNumMV}{\TNum}
  }{MKSNum}

  \judgment{ }{
    \ctxSynTypeM{\ctx}{\ECNumMV}{\TNum}
  }{MSNum}
\end{mathpar}

For addition expressions, the type of both operands should be $\TNum$. To denote this in a
bidirectional system, the operands are analyzed against $\TNum$, giving the following typing rule
for unmarked addition expressions:
%
\begin{mathpar}
  \judgment{
    \ctxAnaTypeU{\ctx}{\EMV_1}{\TNum} \\
    \ctxAnaTypeU{\ctx}{\EMV_2}{\TNum}
  }{
    \ctxSynTypeU{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\TNum}
  }{USPlus}
\end{mathpar}

The marking rule mirrors \textsc{USPlus} closely. Since an expected type for each operand is known,
we shift the responsibility for any type errors to them. Hence, we recursively mark each operand in
analytic mode and rebuild the marked addition expression. The marked typing rule then mirrors
\textsc{USPlus} exactly.
%
\begin{mathpar}
  \judgment{
    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TNum} \\\\
    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TNum}
  }{
    \ctxSynFixedInto{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{MKSPlus}

  \judgment{
    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TNum} \\\\
    \ctxAnaTypeM{\ctx}{\ECMV_2}{\TNum}
  }{
    \ctxSynTypeM{\ctx}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
  }{MSPlus}
\end{mathpar}

Only synthetic rules were necessary for typing numbers and addition because of \emph{subsumption},
given below, which states that if an expression synthesizes a type, it may also be analyzed against
that type or any \emph{consistent} type.
\[%
  \judgment{
    \ctxSynTypeU{\ctx}{\EMV}{\TMV'} \\
    \consistent{\TMV}{\TMV'} \\
    \subsumable{\EMV}
  }{
    \ctxAnaTypeU{\ctx}{\EMV}{\TMV}
  }{UASubsume}
\]%
We rely on the notion of \emph{type consistency} from gradual type theory, which defines a reflexive
and symmetric (but not transitive) relation between types, writing $\consistent{\TMV_1}{\TMV_2}$ to
mean that $\TMV_1$ is consistent with $\TMV_2$. Defined in \cref{fig:calculus-consistency}, this
replaces the notion of \emph{type equality} to relate the unknown type to all other types.

@@

    \judgment{
      \consistent{\TMV_1}{\TMV_1'} \\
      \consistent{\TMV_2}{\TMV_2'} \\
    }{
      \consistent{\TArrow{\TMV_1}{\TMV_2}}{\TArrow{\TMV_1'}{\TMV_2'}}
    }{TCArr} \\
  \end{mathpar}
  %
  \caption{Type consistency.}
  \label{fig:calculus-consistency}
\end{figure}

Hence, when checking $\ENumMV$ against a known expected type, subsumption checks that the type is
consistent with $\TNum$, the type that $\ENumMV$ synthesizes. This succeeds for $\TUnknown$ and
$\TNum$.

We also restrict the usage of subsumption to ``subsumable'' syntactic forms, written
$\subsumable{\EMV}$. This judgment is defined on all syntactic forms except lambda abstractions and
conditionals, the only ones with explicit analysis rules. In other words, we restrict subsumption to
be the rule of ``last resort'', which is necessary to establish that marking and typing are
deterministic (discussed later in \cref{thm:calculus-marking-unicity}).

Now, to define analytic marking on forms without explicit analytic typing rules, we also need
subsumption rules for marking and marked expression typing. Note that we define a analogous notion
of ``subsumability'' for marked expressions, written $\subsumable{\ECMV}$.
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'} \\
    \consistent{\TMV}{\TMV'} \\
    \subsumable{\EMV}
  }{
    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}
  }{MKASubsume}

  \judgment{
    \ctxSynTypeM{\ctx}{\ECMV}{\TMV'} \\
    \consistent{\TMV}{\TMV'} \\
    \subsumable{\ECMV}
  }{
    \ctxAnaTypeM{\ctx}{\ECMV}{\TMV}
  }{MASubsume}
\end{mathpar}

But what happens when the synthesized type of an expression is \emph{not} consistent with the
expected type, i.e. that the premise $\consistent{\TMV}{\TMV'}$ fails? Recalling the example of
\cref{fig:calculus-examples-inconsistent-types}, subsumption would be used to analyze $\ETrue$
against $\TNum$ in $\EPlus{\ETrue}{5}$, which fails since $\inconsistent{\TNum}{\TBool}$, i.e.
\textsc{UASubsume} does not apply, and traditional typing semantics would simply fail the type
checking process at this point. \textsc{MKASubsume} would also not be applicable, leaving marking
undefined in those cases.

% TODO use example more closely related to num here since we haven't talked about booleans yet?

To satisfy marking totality, such a possibility motivates an \emph{inconsistent type mark}
$\ECInconType{\ECMV}$, which states exactly that the synthesized type of $\ECMV$ is inconsistent
with the expected type. It is then accompanied by the following rules:
%
\begin{mathpar}
  \judgment{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'} \\
    \inconsistent{\TMV}{\TMV'} \\
    \subsumable{\EMV}
  }{
    \ctxAnaFixedInto{\ctx}{\EMV}{\ECInconType{\ECMV}}{\TMV}
  }{MKAInconsistentTypes}

  \judgment{
    \ctxSynTypeM{\ctx}{\ECMV}{\TMV'} \\
    \inconsistent{\TMV}{\TMV'} \\
    \subsumable{\ECMV}
  }{
    \ctxAnaTypeM{\ctx}{\ECInconType{\ECMV}}{\TMV}
  }{MAInconsistentTypes}
\end{mathpar}
%
Observe that the premises of \textsc{MKAInconsistentTypes} are identical to those of
\textsc{MKASubsume}, except that $\inconsistent{\TMV}{\TMV'}$. By marking an error, type-checking
may continue---instead of simply failing.

Let us now consider the case of variables. Typing in the unmarked language is standard, and because
of subsumption, only a synthetic rule is required:
\[%
  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynTypeU{\ctx}{x}{\TMV}
  }{USVar}
\]%
That is, if $x$ is bound to a type in the typing context, it synthesizes that type.
Straightforwardly, marking converts an unmarked variable into the same variable via the following
rules:
%
\begin{mathpar}
  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{x}{x}{\TMV}
  }{MKSVar}

  \judgment{
    \inCtx{\ctx}{x}{\TMV}
  }{
    \ctxSynTypeM{\ctx}{x}{\TMV}
  }{MSVar}
\end{mathpar}
% TODO more intuition on why this makes sense?

However, consider the case that a variable, such as $y$ in \cref{fig:calculus-examples-unbound}, is \emph{not}
bound. Again, \textsc{USVar} would not apply. A total marking should, however, report the error and
continue, motivating an \emph{unbound variable mark} $\ECUnbound{x}$ and the accompanying rules:
%
\begin{mathpar}
  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynFixedInto{\ctx}{x}{\ECUnbound{x}}{\TUnknown}
  }{MKSUnbound}

  \judgment{
    \notInCtx{\ctx}{x}
  }{
    \ctxSynTypeM{\ctx}{\ECUnbound{x}}{\TUnknown}
  }{MSUnbound}
\end{mathpar}
%
An unbound variable is marked as such, and since nothing may be said about their types, we
synthesize the unknown type. Similarly to the inconsistent type mark, this allows type checking to
proceed, with usage of the unbound variable permitted in any expression.
% TODO clarify/give intuition as to why this is correct

Unlike numbers and variables, there are explicit synthesis and analysis rules for unmarked lambda
abstractions: % TODO intuition why
%
\begin{mathpar}
  \judgment{
    \ctxSynTypeU{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxSynTypeU{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{USLam}

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\
    \ctxAnaTypeU{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxAnaTypeU{\ctx}{\ECLam{x}{\TMV}{\EMV}}{\TMV_3}
  }{UALam}
\end{mathpar}
%
The synthesis rule is standard. The analysis rule employs the judgment
$\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$, which establishes that $\TMV$ is a \emph{matched arrow type}
[citation], i.e. it may be considered an arrow type. Defined in \cref{fig:calculus-matched-arrow},
this notion is purely a technical mechanism to avoid duplication of rules related to arrow types.

\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\matchedArrow{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched arrow type $\TArrow{\TMV_1}{\TMV_2}$
  %
  \begin{mathpar}
    \judgment{ }{
      \matchedArrow{\TUnknown}{\TUnknown}{\TUnknown}
    }{TMAHole}

    \judgment{ }{
      \matchedArrow{\TArrow{\TMV}{\TMV}}{\TMV}{\TMV}
    }{TMAArr}
  \end{mathpar}
  %
  \caption{Matched arrow types.}
  \label{fig:calculus-matched-arrow}
\end{figure}

@@

  \judgment{
    \ctxSynTypeM{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
  }{
    \ctxSynTypeM{\ctx}{\ECLam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{MSLam}
\end{mathpar}
%
Similarly, we construct corresponding analytic rules:
%
\begin{mathpar}
  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\\\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MKALam1}

  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\\\
    \ctxAnaTypeM{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
  }{
    \ctxAnaTypeM{\ctx}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MALam1}
\end{mathpar}
%
However, recalling totality, we look again to the premises of \textsc{UALam} to see what type errors
may arise. First, what if $\TMV_3$ is not a matched arrow type? This would be the case if it were
$\TNum$, for example. The lambda abstraction's synthesized type would be inconsistent with the
expected type, but this case is slightly distinct from that of the inconsistent types mark;

Thus,
to the syntax we add an \emph{inconsistent type mark} $\ECInconType{\ECMV}$, which indicates that
the expression does not synthesize a type consistent with the expected type. We then have the
following rules, in which the lambda abstraction is placed inside the inconsistent type mark.
%
\begin{mathpar}
  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TUnknown}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLamAnaNonMatchedArrow{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MKALam2}

  \judgment{
    \notMatchedArrow{\TMV_3} \\
    \ctxAnaTypeM{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TUnknown}
  }{
    \ctxAnaTypeM{\ctx}{\ECLamAnaNonMatchedArrow{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MALam2}
\end{mathpar}
%
Though no expected output type is known, we still need to check and mark the body; it is thus
analyzed against the unknown type in \textsc{MKALam2}.
